Insert your title here

### 2018-08-11 sect3-exact-methods.tex
Decision version of the binomial tree model was stated with objective function maximizing number of assigned labels. Another simple upper bound based on Prim's alg. was described.

%In Model \ref{mod:basic}, superscript of $x$-variables represents a time step in which a signal is transmitted along corresponding arc.
The nature of MBT suggests another modelling approach derived from Model \ref{mod:basic}. 
For a given maximum broadcast time $t_{\text{max}}$, we maximize the number of arcs along which a signal is relayed within $t_{\text{max}}$.
If the optimal value is $n-1$, it is clear that we found a spanning broadcast tree with broadcast time $t_{\text{max}}$.
In case the optimal value is less than $n-1$, the gien $t_{\text{max}}$ is insufficient and we have to try solving the problem with increased time limit.
Clearly, an additional knowledge of upper and lower bound spares some computations. 
A lower bound is the initial value of $t_{\text{max}}$. 
Similarly, if an upper bound $\bar{t}$ is known and $t_{\text{max}}=\bar{t}-1$, the process can be terminated when the objective function is less then $n-1$, because $\bar{t}$ is than the optimal value.

@@

\subsection{Binomial tree model}

@@

>>>>

%In Model \ref{mod:basic}, superscript of $x$-variables represents a time step in which a signal is transmitted along corresponding arc.
\subsubsection{Decision version}
\label{sec:decbasic}
The nature of MBT suggests another modelling approach derived from Model \ref{mod:basic}. 
For a given maximum broadcast time $t_{\text{max}}$, we maximize the number of arcs along which a signal is relayed within $t_{\text{max}}$.
If the optimal value is $n-1$, it is clear that we found a spanning broadcast tree with broadcast time $t_{\text{max}}$.
In case the optimal value is less than $n-1$, the gien $t_{\text{max}}$ is insufficient and we have to try solving the problem with increased time limit.
Clearly, an additional knowledge of upper and lower bound spares some computations. 
A lower bound is the initial value of $t_{\text{max}}$. 
Similarly, if an upper bound $\bar{t}$ is known and $t_{\text{max}}=\bar{t}-1$, the process can be terminated when the objective function is less then $n-1$, because $\bar{t}$ is than the optimal value.

@@

The objective function \eqref{mod:basic:dec:obj} maximizes the number of arcs relaying a signal.
The actual target function is the minimum time required for spreading the signal to all nodes, and so we are looking for the smallest value of $t_{text{max}}$.
\subsection{Binomial tree model}

@@

\subsubsection{Decision version}

In a similar manner as we derived the decision version \eqref{mod:basic:dec} of model \eqref{mod:basic}, it is possible to construct a decision version of the binomial tree model \eqref{mod:partition} as follows:
\begin{subequations}\label{mod:genmatch}
\begin{align}
\notag \max\sum\limits_{v\in V}&\sum\limits_{i\in I}\sum\limits_{s\in S}   y_{is}^v,\\
\notag \text{s. t. } \\
\label{mod:genmatch:nodeBelongs} \sum\limits_{i\in I}\sum\limits_{s\in S}y^v_{is} & \leq 1 & v\in V,\\
\notag\eqref{mod:part:treeHasIJ} - \eqref{mod:part:followArcsB},\\
%\label{mod:genmatch:treeHasIJ} \sum\limits_{v\in V'}y^v_{is} & = 1 & i\in I,s\in V_T,\\
%\label{mod:genmatch:source1} y_{1s}^s & = 1  & s\in V_T,\\
%\label{mod:part:noReturn} y^u_{ij}+y^v_{lj} &\leq 1 & i\in I,l\in C(i), j\in J, u\in V_\alpha,v\in V,\\
%\label{mod:part:followArcs} y^u_{is}+y^v_{\ell s} &\leq 1 & i\in I,\ell\in C(i), s\in S, u,v\in V',(u,v)\not\in A',\\
%\label{mod:part:followArcsA} y^u_{is}+y^v_{\ell s} &\leq 1 & i\in I,\ell\in C(i), s\in S, u,v\in V,\{u,v\}\not\in E,\\
%\label{mod:part:followArcsB} y^{v_0}_{is}+y^v_{\ell s} &\leq 1 & i\in I,\ell\in C(i), s\in S, v\in V,\\
%\label{mod:genmatch:followArcsA}y^{v_0}_{is}+y^u_{i s} + \sum\limits_{v\in V\setminus N(u)}y^v_{\ell s}&\leq 1 & u\in V,i\in I,\ell\in C(i), s\in V_T,  \\
%\label{mod:genmatch:followArcsB}y^{v_0}_{is}+y^u_{\ell s} + \sum\limits_{v\in V\setminus N(u)}y^v_{i s}&\leq 1 & u\in V,i\in I,\ell\in C(i), s\in V_T,\\ 
\label{mod:genmatch:dim}&&y \in \{0,1\}^{I\times S\times V'}.
\end{align}~
\end{subequations}
This model is a modification of formulation \eqref{mod:partition}, and uses the same type of variables.
The objective function is to maximize the number of nodes involved in the binomial trees.
In order to determine the minimum broadcast time, we are looking for the minimum $k$ such that each node can be assigned a label from the set $I=\{1,\dots,2^k\}$ while satisfying given constraints.
%The index set $I=\{1,\dots,2^k\}$ depends on the input parameter $k$.
The constraints \eqref{mod:genmatch:nodeBelongs} state that each node belongs to at most one binomial tree.
Compared to \eqref{mod:part:nodeBelongs}, \eqref{mod:genmatch:nodeBelongs} is an inequality, because the binomial trees do not necessarily form a partition of $G$, and so not all nodes have to be used.
The remaining constraints are taken from formulation \eqref{mod:partition}.

The approach is then analogous to the idea indicated in Section \ref{sec:decbasic}.
Assume we are given a lower bound $\underline{t}$ and an upper bound $\bar{t}$ on the minimum broadcast time.
Initially, we define the set $I=\{1,\dots,2^{\underline{t}}\}$ and iteratively solve the model while doubling the set $I$ until the objective function attains the value $n$.
That indicates that it is the first iteration in which all nodes are assigned a label, and it can be concluded that the minimum broadcast time is $\log_2|I|$.

### 2018-08-11 sect5-upper-bounds.tex
Decision version of the binomial tree model was stated with objective function maximizing number of assigned labels. Another simple upper bound based on Prim's alg. was described.

%The algorithms presented in this section iteratively construct broadcast foerest $T_i, i=1,2,\dots$.
The following inexact method is based on the idea of finding maximum cardinality matching in $G\left[V_{T_i},V\setminus V_{T_i}\right]$, and extending $T_i$ by this matching.
It means that in each iteration $i$, the maximum possible number of nodes are informed.
The maximum cardinality matching can be regarded as finding $|V_{T_i}|$ node-disjoint  binomial trees of order at most one with roots in $V_{T_i}$, maximizing the number of edges.
By generalizing this notion, we iteratively look for $|V_{T_i}|$ node-disjoint (pruned) binomial trees of an arbitrary given order $k$ valued between 1 and $n-|S|$.
Even though this problem is NP-hard for $k\geq 2$ \cite{jansen95}, it is expected that the computational time is sound in most practical instances.
After obtaining the set of binomial trees, first $p$ nodes in each tree are selected and added to the broadcast forest.
The parameter $p\in \{1,\dots,2^k\}$ is also a part of the input.
\begin{algorithm}[]
\KwData{$G=(V,E), S\subseteq V, k\in \{1,\dots,n-|S|\}, p\in \{1,\dots,2^k\}$}
\textbf{for }$s\in S\textbf{ do }T_s=(V_s,A_s), V_s\leftarrow \{s\}, E_s\leftarrow\emptyset$\;
$I=\{1,\dots,2^k\}$\\
$T=\{T_s:s\in S\}$\ ~~~\tcp{broadcast forest}
$t\leftarrow 0$\;
\While{$V_T\neq V$} {
	$S\leftarrow V_T$\;
	Find a set of pruned binomial trees $B=\{B_1,\dots,B_{|V_T|}\}$ of order at most $k$ with roots in $V_T$ by solving model \eqref{mod:genmatch}\;
	$V_T\leftarrow V_T\cup \{v:v\in V_B:\beta(v)\leq p\}$\;
	$E_T\leftarrow E_T\cup \{\{u,v\}\in E_B: \beta(u)\leq p,\beta(v)\leq p\}$\;
	$t\leftarrow t+1$\;
}
\Return t\;
%\Return $\lceil k/s \rceil$\;
 \caption{A method for determining an upper bound}
\label{alg:match}
\end{algorithm}

Alg. \ref{alg:match} describes the process formally.
Initially, the broadcast forest consists of isolated sources.
The binomial trees are determined by solving the ILP model \eqref{mod:genmatch}. 
This model is a modification of formulation \eqref{mod:partition}, and uses the same type of variables.
The objective function is to maximize the number of nodes involved in the binomial trees.
%The index set $I=\{1,\dots,2^k\}$ depends on the input parameter $k$.
The constraints \eqref{mod:genmatch:nodeBelongs} state that each node belongs to at most one binomial tree.
Compared to \eqref{mod:part:nodeBelongs}, \eqref{mod:genmatch:nodeBelongs} is an inequality, because the binomial trees do not necessarily form a partition of $G$, and so not all nodes have to be used.
The remaining constraints are taken from formulation \eqref{mod:partition}. 
Note that in every iteration, the model is solved with a different set $S$, as all nodes already included in the broadcast forest are roots of the binomial trees.
The model considers the entire set $V$, but it is also possible to restrict this set to nodes with distance at most $k$ from some node in $V_T$.
When this restriction is not imposed, nodes with larger distance are not a part of any binomial tree in the current iteration due to \eqref{mod:part:followArcsA} - \eqref{mod:part:followArcsB}.

\begin{subequations}\label{mod:genmatch}
\begin{align}
\notag \max\sum\limits_{v\in V}&\sum\limits_{i\in I}\sum\limits_{s\in S}   y_{is}^v,\\
\notag \text{s. t. } \\
\label{mod:genmatch:nodeBelongs} \sum\limits_{i\in I}\sum\limits_{s\in S}y^v_{is} & \leq 1 & v\in V,\\
\notag\eqref{mod:part:treeHasIJ} - \eqref{mod:part:followArcsB},\\
%\label{mod:genmatch:treeHasIJ} \sum\limits_{v\in V'}y^v_{is} & = 1 & i\in I,s\in V_T,\\
%\label{mod:genmatch:source1} y_{1s}^s & = 1  & s\in V_T,\\
%\label{mod:part:noReturn} y^u_{ij}+y^v_{lj} &\leq 1 & i\in I,l\in C(i), j\in J, u\in V_\alpha,v\in V,\\
%\label{mod:part:followArcs} y^u_{is}+y^v_{\ell s} &\leq 1 & i\in I,\ell\in C(i), s\in S, u,v\in V',(u,v)\not\in A',\\
%\label{mod:part:followArcsA} y^u_{is}+y^v_{\ell s} &\leq 1 & i\in I,\ell\in C(i), s\in S, u,v\in V,\{u,v\}\not\in E,\\
%\label{mod:part:followArcsB} y^{v_0}_{is}+y^v_{\ell s} &\leq 1 & i\in I,\ell\in C(i), s\in S, v\in V,\\
%\label{mod:genmatch:followArcsA}y^{v_0}_{is}+y^u_{i s} + \sum\limits_{v\in V\setminus N(u)}y^v_{\ell s}&\leq 1 & u\in V,i\in I,\ell\in C(i), s\in V_T,  \\
%\label{mod:genmatch:followArcsB}y^{v_0}_{is}+y^u_{\ell s} + \sum\limits_{v\in V\setminus N(u)}y^v_{i s}&\leq 1 & u\in V,i\in I,\ell\in C(i), s\in V_T,\\ 
\label{mod:genmatch:dim}&&y \in \{0,1\}^{I\times S\times V'}.
\end{align}~
\end{subequations}

>>>>

A knowledge of an upper bound $\bar{t}$ affects the number of variables in all studied models. 
Particularly in the decision versions, the iterative approach can be terminated once the solution is dissatisfactory for broadcast time limit $\bar{t}-1$.
The algorithms presented in this section iteratively construct broadcast foerest $T_i, i=1,2,\dots$.
\subsection{Prim based method}

A fast straightforward algorithm determining an upper bound is inspired by Prim algorithm for minimum spanning tree problem.
Initially, the constructed broadcast forest consists of all sources. 
In each iteration, the nodes already in the forest are sorted according to the number of neighbors not yet added to the forest.
For each node we then select a neighbor to be appended to the forest.
\begin{algorithm}[]
\KwData{$G=(V,E), S\subseteq V$}
\textbf{for }$s\in S\textbf{ do }T_s=(V_s,A_s), V_s\leftarrow \{s\}, E_s\leftarrow\emptyset$\;
$T=\{T_s:s\in S\}$\ ~~~\tcp{broadcast forest}
$t\leftarrow 0$\;
\While{$V_T\neq V$} {
	Sort nodes in $V_T$ increasingly by the number of neighbors in $V\setminus V_T$\;
	$W\leftarrow\emptyset$\;
	\For{$u\in V_T$} {
		Find $u\in V\setminus (V_T\cup W)$ according to some strategy\; 	
	}
	$E_T\leftarrow E_T\cup \{\{u,w\}: u\in V_T, w\in N(u)\cap W\}$\;
	$V_T\leftarrow V_T\cup W$\;
	$t\leftarrow t+1$\;
}
\Return t\;
%\Return $\lceil k/s \rceil$\;
 \caption{A greedy Prim-like method for determining an upper bound}
\label{alg:match}
\end{algorithm}

\subsection{Restricted binomial tree method}

The following inexact method is based on the idea of finding maximum cardinality matching in $G\left[V_{T_i},V\setminus V_{T_i}\right]$, and extending $T_i$ by this matching.
It means that in each iteration $i$, the maximum possible number of nodes are informed.
The maximum cardinality matching can be regarded as finding $|V_{T_i}|$ node-disjoint  binomial trees of order at most one with roots in $V_{T_i}$, maximizing the number of edges.
By generalizing this notion, we iteratively look for $|V_{T_i}|$ node-disjoint (pruned) binomial trees of an arbitrary given order $k$ valued between 1 and $n-|S|$.
Even though this problem is NP-hard for $k\geq 2$ \cite{jansen95}, it is expected that the computational time is sound in most practical instances.
After obtaining the set of binomial trees, first $p$ nodes in each tree are selected and added to the broadcast forest.
The parameter $p\in \{1,\dots,2^k\}$ is also a part of the input.
\begin{algorithm}[]
\KwData{$G=(V,E), S\subseteq V, k\in \{1,\dots,n-|S|\}, p\in \{1,\dots,2^k\}$}
\textbf{for }$s\in S\textbf{ do }T_s=(V_s,A_s), V_s\leftarrow \{s\}, E_s\leftarrow\emptyset$\;
$I=\{1,\dots,2^k\}$\\
$T=\{T_s:s\in S\}$\;
$t\leftarrow 0$\;
\While{$V_T\neq V$} {
	$S\leftarrow V_T$\;
	Find a set of pruned binomial trees $B=\{B_1,\dots,B_{|V_T|}\}$ of order at most $k$ with roots in $V_T$ by solving model \eqref{mod:genmatch}\;
	$V_T\leftarrow V_T\cup \{v:v\in V_B:\beta(v)\leq p\}$\;
	$E_T\leftarrow E_T\cup \{\{u,v\}\in E_B: \beta(u)\leq p,\beta(v)\leq p\}$\;
	$t\leftarrow t+1$\;
}
\Return t\;
%\Return $\lceil k/s \rceil$\;
 \caption{A method for determining an upper bound based on iterative search for pruned binomial trees}
\label{alg:match}
\end{algorithm}

Alg. \ref{alg:match} describes the process formally.
Initially, the broadcast forest consists of isolated sources.
The binomial trees are determined by solving the ILP model \eqref{mod:genmatch}. 
Note that in every iteration, the model is solved with a different set $S$, as all nodes already included in the broadcast forest are roots of the binomial trees.
The model considers the entire set $V$, but it is also possible to restrict this set to nodes with distance at most $k$ from some node in $V_T$.
When this restriction is not imposed, nodes with larger distance are not a part of any binomial tree in the current iteration due to \eqref{mod:part:followArcsA} - \eqref{mod:part:followArcsB}.

Not only determine the algorithms the upper bound on the minimum broadcast time, but they also construct a solution with this broadcast time.
